---
title: "SCOscore: School Corporation Opportunity Score"
author:
  - name: Jeremy Price 
    url: https://www.jeremyfprice.info/
    affiliation: CEISL / IU School of Education-Indianapolis
    affiliation_url: https://education.iupui.edu/
    orcid: 0000-0002-6506-3526
  - name: Akaash Arora
    affiliation: CEISL / IU School of Education-Indianapolis
    affiliation_url: https://education.iupui.edu/
  - name: Maxim Bulanov
    affiliation: CEISL / IU School of Education-Indianapolis
    affiliation_url: https://education.iupui.edu/
  - name: AJ Knoors
    affiliation: CEISL / IU School of Education-Indianapolis
    affiliation_url: https://education.iupui.edu/
license: "CC BY-SA"
code-annotations: true
mainfont: ETbb
sansfont: "Gill Sans"
monofont: "JetBrains Mono"
format: pdf
highlight-style: a11y
code-overflow: wrap
reference-location: margin
cap-location: margin
tbl-cap-location: margin
link-external-icon: false
link-external-newwindow: true
repo-url: https://github.com/jeremyfprice/scoscore
citation:
  type: document
  issued: 2023
  url: https://github.com/jeremyfprice/scoscore
  doi: 10.17605/OSF.IO/ZT9UX
google-scholar: true
bibliography: scoscores-references.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r load-libraries}
library(readr)
library(tidyr)
library(dplyr)
library(here)
library(DT)
library(ggplot2)
library(ggpubr)
library(plotly)
library(ggbeeswarm)
library(edbuildmapr)
library(maps)
library(mapdata)
library(ggmap)
#library(akima)
library(ggrepel)
library(sf)

ucl_palette <- c(
  "City large" = "#990000",
  "City midsize" = "#F23A3F",
  "City small" = "#FF636A",
  "Suburb large" = "#FFAA00",
  "Suburb midsize" = "#FFD563",
  "Suburb small" = "#FFE694",
  "Town fringe" = "#056E41",
  "Town distant" = "#63B363",
  "Town remote" = "#A7D094",
  "Rural fringe" = "#006298",
  "Rural distant" = "#63B1D3",
  "Rural remote" = "#94D2E7"
)
```
```{r read-data}
SCO_frame <- read_csv(here("data", "in_scoscores.csv"), show_col_types = FALSE) |>
  na.omit()

SCO_frame$urm_pct <- round(SCO_frame$urm_pct, digits = 2)
SCO_frame$frl_pct = round(SCO_frame$frl_pct, digits = 2)
SCO_frame$academic = round(SCO_frame$academic, digits = 2)
SCO_frame$scoScore = round(SCO_frame$scoScore, digits = 2)
```

## Introduction

[This is an Open Access Project {{< ai open-access title='open access' >}}]{.aside}

The School Corporation Opportunity Score (SCOscore) is a measure of potentiality for learners in a
school community. It is a composite score that combines multiple factors, particularly structural
factors such as race and SES, as well as performance factors such as test scores and graduation
pathway completion rates, to provide a more holistic view of school corporations.

## Methods

The SCOscore of a school corporation is calculated through the following formula:

$$
S_{O} = \frac{\left[\left(P_{urm} \times 1.5\right) + \left(P_{frl} \times 1.5\right) + S_{ac}\right]}{3}
$$

where $S_{O}$ is the SCOscore, $P_{urm}$ is the proportion of underrepresented minority students,
$P_{frl}$ is the proportion of students eligible for free or reduced lunch, and $S_{ac}$ is an academic
achievement score for the corporation as a whole.

The academic achievement score accounts for 3rd grade ELA proficiency, 8th grade 
math proficiency, and graduation pathways completion rates. In order to account
for those factors, the following equation is utilized:

$$
S_{ac} = \frac{\left(S_{a_{ela}}^\prime + S_{a_{math}}^\prime + S_{a_{gpc}}^\prime\right)}{3} + 1
$$
where $S_{a_{ela}}^\prime$ is the calculated 3rd grade ELA proficiency measure,
$S_{a_{math}}^\prime$ is the calculated 6th grade math proficiency measure, and
$S_{a_{gpc}}^\prime$ is the calculated graduation pathway completion measure.

The $S_a^\prime$ measures are determined in the following manner:

$$
f(S_a^\prime) = \begin{cases}
(S_{IN} - S_a) + 1, & \text{if } S_a < S_{IN} \\
0, &\text{otherwise}
\end{cases}
$$

where $S_a$ is the school corporation average proficiency or completion rate and 
$S_{IN}$ is the state-level average proficiency or completion rate.

### Acknowledging Inequality *Within* School Corporations

It needs to be recognized that there is the potential for inequality *within* a
single school corporation and the SCOscore is calculated based on reported *averages*
across all schools in the corporation. The school corporation-level score should be considered
within the local building context. A building or school-level score can be calculated using
the same general equation:

$$
S_{O} = \frac{\left[\left(P_{urm} \times 1.5\right) + \left(P_{frl} \times 1.5\right) + S_{ac}\right]}{3}
$$

Rather than calculating a combined academic score, however, the specific academic
measure--3rd grade ELA proficiency rates for elementary schools, 6th grade math
proficiency rates for middle schools, and graduation pathways completion rates for
high schools--can be used. The $S_a^\prime$ still needs to be calculated by
comparing the measure to the Indiana statewide average as above, where it becomes
$0$ if it is above the statewide average or $(S_{IN} - S_a) + 1$ if it is below
the statewide average.

::: {.callout-important}
## Charter Schools
It is also important to consider the status of charter schools when identifying
schools to work with. Despite the ["public school" designation](https://www.in.gov/icsb/families-and-students/charter-school-faqs/)
by the Indiana Department of Education, [charter schools are not technically public schools](https://www.forbes.com/sites/petergreene/2019/02/02/charter-schools-are-not-public-schools/?sh=17e935536832) in important ways, particularly when it comes to accountability to the community
in which they are embedded. Other problematic issues with respect to oppression by
[race](https://journals.sagepub.com/doi/full/10.1177/1478210319875385) and
[ability](https://dsq-sds.org/article/%20view/3187/3072) are well-documented.
:::

### SCO Scores and Critical Quantitative Methodologies

The School Corporation Opportunity Score model can be seen as *consistent* with
the five tenents of Critical Quantitative (QuantCrit) methodologies [@gillbornQuantCritEducationPolicy2018],
or at least can be utilized in light of QuantCrit. In particular, the SCO scores model
is *intended* to be utilized for social justice, recognizes the centrality of racism,
pushes back on static categories, and through transparency attempts to illuminate
that numbers are not neutral.

We also recognize, however, that the SCO scores model *violates* several principal
values of the QuantCrit methodology [@youngDecodingDataDichotomy2022]. SCO scores
are based on comparative assumptions, comparing typically "successful" school corporations
(predominantly white, predominantly wealthy) with those that are not. Even with
an assets-based framing ("opportunity" rather than "need")--or at least a positive
spin--school corporations that are predominantly white and wealthy "[hover] over
Black and Latinx scores, which creates academic imagery that promotes racial
achievement hierarchies and White supremacy" [@youngDecodingDataDichotomy2022, p. 392].
Even though it is the school corporations with high Black, brown, immigrant,
refugee, and poor populations that rise up, it still creates a sense of (inverted)
hierarchies.

What would an approach more tightly entwined with QuantCrit methodologies look like?
The *single-group summary meta-analysis* proposed by Young and Young [-@youngDecodingDataDichotomy2022]
provides guidance. Academic achievement--through the measures provided by the
Indiana Department of Education's GPS Dashboard [@IndianaGraduatesPrepared]--can be examined over time for
specific groups of interest (e.g., Black students, Latinx students, refugee students,
recipients of free or reduced lunch, etc.) and an effect size can be calculated for
each group independently. If the effect size for *any* particular group is less than
or equal to 0.4, that school corporation becomes a high opportunity school corporation
because the potential achievement of that group of students has not yet been realized.
This would provide a much more textured and measured approach to understanding the
experiences of under- or mis-supported student populations.

### Data Sources

Data for these calculations are drawn from the following sources:

* Indiana Department of Education's [Indiana Graduates Prepared to Succeed (GPS)](https://indianagps.doe.in.gov/) Dashboard [@IndianaGraduatesPrepared].
* Urban Institute's [Education Data Explorer](https://educationdata.urban.org/data-explorer) via
the educationdata R package [@ueyamaEducationdata].

## School Corporation Opportunity Scores (SCO Scores)

This section will highlight three areas for consideration: School corporation SCO scores
in central Indiana; SCO score trends across the state; and, a brief examination of
SCO scores by urban-centric locale categories.

There is an [Indiana SCO Score Dashboard](https://jeremyfprice.github.io/SCOscore-dashboard/) available, in which you can explore
the data in more detail and depth.

```{r sc-map, message = FALSE, warning=FALSE, output = FALSE}
states <- sf::st_as_sf(map("state", region = "indiana", plot = FALSE, fill = TRUE))

in_sc_shapes <- sd_shapepull("2019", with_data = TRUE) |>
  filter(State == "Indiana") |>
  select(leaid = GEOID, geometry) |>
  mutate(leaid = as.numeric(leaid))

SCO_frame <- SCO_frame |>
  right_join(in_sc_shapes)

school_corp_map <- st_as_sf(SCO_frame)

# corp_map <- SCO_frame |>
#   select(leaid, scoScore, geometry) |>
#   st_as_sf(map("state", plot = FALSE, fill = TRUE))

sco_map <- ggplot(data = states) +
  geom_sf(data = states, fill = "#eeeeee") +
  geom_sf(data = school_corp_map, color = "#243142", aes(
    label = lea_name,
    fill = scoScore)) +
  scale_fill_steps2(
    name = "SCOscore",
    low = "#FFF4C6",
    high = "#990000",
    mid = "#FFAA00",
    midpoint = 1.1,
    breaks = c(0, 0.6, 0.8, 1, 1.2, 1.4)
  ) +
  coord_sf(xlim = c(-88.5, -84.5), ylim = c(37.5, 42), expand = FALSE) +
  # annotate("text", x = -89.25, y = 40, label = "ILLINOIS") +
  # annotate("text", x = -83.25, y = 40, label = "OHIO") +
  # annotate("text", x = -84.75, y = 38, label = "KENTUCKY") +
  theme_minimal()  +
  theme(panel.grid.major = element_blank(),
        # panel.background = element_rect(fill = "#E9F6FC"),
        axis.text = element_blank(),
        axis.title = element_blank())
ggsave(here("outputs", "indiana_wo_campuses.png"), sco_map, dpi = 300, width = 9, height = 16, units = "in")
```

### Central Indiana

The primary purpose of this model is to provide insight into identifying high-opportunity
school corporations for IU-Indianapolis faculty, staff, and projects. The school corporations
in Central Indiana (Indianapolis–Carmel–Anderson Metropolitan Statistical Area)
with SCOscores greater than 1 are as follows:

```{r display-table}
#| label: tbl-central-indiana
#| tbl-cap: School corporations in the Indianapolis–Carmel–Anderson Metropolitan Statistical Area with SCO scores greater than 1.

table_frame <- data.frame(
  "School Corporation" = c(
    "Indianapolis Public Schools",
    "MSD Warren Twp",
    "MSD Wayne Twp",
    "MSD Pike Twp",
    "Anderson Community Schools",
    "MSD Lawrence Twp",
    "MSD Washington Twp",
    "Perry Twp Schools",
    "Beech Grove City Schools",
    "MSD Decatur Twp"
  ),
  Enrollment = c(
    22115,
    11801,
    16343,
    10928,
    6470,
    16247,
    10901,
    16603,
    2837,
    6681
  ),
  "SCO Score" = c(
    1.42,
    1.42,
    1.39,
    1.37,
    1.30,
    1.29,
    1.13,
    1.11,
    1.10,
    1.07
  )
)

library(knitr)
kable(table_frame)
```

```{r central-indiana-map}
#| label: fig-central-indiana-map
#| fig-cap: "A map of school corporations in the Indianapolis–Carmel–Anderson Metropolitan Statistical Area with SCO scores."
#| column: margin

central_indiana_map <- sco_map +
  coord_sf(xlim = c(-86.5, -85.45), ylim = c(39.5, 40.25), expand = FALSE) +
  theme(legend.position = "top")

central_indiana_map
```

Most of the high opportunity schools are centered in Marion County (with the notable
exception of Anderson Community Schools). Indianapolis Public schools and the Metropolitan
School District (MSD) of Warren Township are the school corporations with the *highest
opportunity scores* in central Indiana, with a SCO score of 1.42. The MSD of Wayne
Township and the MSD of Pike Township also have high opportunity scores (1.39 and 1.37,
respectively). The other school corporations listed are also high opportunity.
**Please note the sections above on inequalities within school corporations and the
status of charter schools when reaching out to a specific school building.**

### State Trends

Overall, most of Indiana's school corporations' opportunity score fall under 1 (@fig-density-plot).
This indicates that the majority of Indiana school corporations have low levels 
of non-white students, low levels of students who qualify for free or reduced lunch,
relatively high achievement, or some combination of the three. Many of the school
corporations in Indiana fall between the 0.5 and 1 range. This doesn't mean that
there *isn't* opportunity in these school corporations, particularly for those
students who experience structure barriers and oppression; rather, as a matter of
general priorities, the school corporations that score greater than 1 will see
a higher level of potential in terms of growth.

```{r density-plot}
#| label: fig-density-plot
#| fig-cap: "A density plot of SCO scores across the state of Indiana."

density_plot <- ggdensity(SCO_frame, x = "scoScore", fill = "#990000", rug = TRUE) +
  scale_x_continuous(limits = c(0, 2)) +
  theme_pubr()
density_plot
```

Over the state of Indiana as a whole (@fig-indiana-map), high opportunity school corporations are centered
in Marion County (the Indianapolis area) and Lake County (the Northwest Region of
Indiana). There are certainly other school corporations around the state that score
above 1 in terms of opportunity, but they do not cluster in the same way as those
in Marion and Lake Counties.

```{r hi-res-map}
#| label: fig-indiana-map
#| fig-cap: "A map of school corporations in Indiana with SCO scores."

cities_frame <- data.frame(
  city = c(
    "IU Indianapolis",
    "IU Northwest",
    "IU South Bend",
    "IU Ft Wayne",
    "IU East",
    "IU Bloomington",
    "IU Kokomo",
    "IU South East",
    "IUPUC"
  ),
  latitude = c(
    39.77733,
    41.59393,
    41.67453,
    41.07438,
    39.8245,
    39.14494,
    40.4768618,
    38.3147169,
    39.1977044
  ),
  longitude = c(
    -86.15460,
    -87.40824,
    -86.24909,
    -85.13782,
    -84.90206,
    -86.53058,
    -86.2202796,
    -85.9060672,
    -86.0643281
  )
)

hi_res_map <- sco_map +
  geom_point(data = cities_frame, aes(x = longitude, y = latitude), color = "#243142", size = 2) +
   geom_text_repel(data = cities_frame, aes(x = longitude, y = latitude, label = city), color = "#243142",
                    nudge_x = c(-2, -1, 1, 1.25, 1, -2), nudge_y = c(0.25, -0.25, 0.25, -0.5, -0.5))
ggsave(here("outputs", "indiana_w_campuses.png"), hi_res_map, dpi = 300, width = 9, height = 16, units = "in")
hi_res_map
```

### Urban-Centric Locale SCO Scores

The National Center for Educational Statistics utilizes the [urban-centered locale](https://nces.ed.gov/pubs2007/ruraled/exhibit_a.asp)
classification to categorize school districts and corporations into rural, town,
suburban, and urban areas. These categories are further divided into large, medium,
and small for urban and suburban areas and fringe, distant, and remote for towns and
rural areas.

::: {.callout-note}
School corporations in Indiana span the full spectrum of these categories.
While there is a higher *number of school corporations* that are rural, a much larger
*number of students* attend urban and suburban schools.
:::

```{r}
#| label: fig-donuts
#| fig-cap: "Total enrollment and number of school corporations by urban-centered locale classification."

urcl_frame <- SCO_frame |>
  group_by(urban_centric_locale) |>
  summarize(num = n(),
            total_enrollment = sum(enrollment)) |>
  na.omit()

urcl_frame$urban_centric_locale <- factor(
  urcl_frame$urban_centric_locale,
  levels = c(
    "City large",
    "City midsize",
    "City small",
    "Suburb large",
    "Suburb midsize",
    "Suburb small",
    "Town fringe",
    "Town distant",
    "Town remote",
    "Rural fringe",
    "Rural distant",
    "Rural remote"
  )
)

urcl_enrollment <- ggdonutchart(
  urcl_frame,
  "total_enrollment",
  label = "urban_centric_locale",
  fill = "urban_centric_locale",
  font.label = list(size = 1),
  palette = ucl_palette) +
  theme(legend.position = "none")
# urcl_enrollment
# ggsave(here("outputs", "enrollment_donut.png"), urcl_enrollment, dpi = 300, width = 8.5, height = 11, units = "in")

urcl_n <- ggdonutchart(
  urcl_frame,
  "num",
  label = "urban_centric_locale",
  fill = "urban_centric_locale",
  palette = ucl_palette) +
  theme(legend.position = "none")

# urcl_n
# ggsave(here("outputs", "scn_donut.png"), urcl_n, dpi = 300, width = 8.5, height = 11, units = "in")

donut_charts <- ggarrange(urcl_enrollment, urcl_n, ncol = 2)
donut_charts
```

@fig-beeswarm visualizes the SCO scores by urban-centered locale classification.
Each of the dots represents an individual school corporation. This plot provides
insight into how the school corporation opportunity scores are distributed across
the urban-centered local classifications.

```{r plot-beeswarm, results = "asis"}
#| label: fig-beeswarm
#| fig-cap: "Plot of SCO scores by urban-centered locale classification."

SCO_frame$urban_centric_locale <- factor(
  SCO_frame$urban_centric_locale,
  levels = c(
    "City large",
    "City midsize",
    "City small",
    "Suburb large",
    "Suburb midsize",
    "Suburb small",
    "Town fringe",
    "Town distant",
    "Town remote",
    "Rural fringe",
    "Rural distant",
    "Rural remote"
  )
)
SCO_frame <- SCO_frame |>
  na.omit()
urcl_swarm <- ggplot(
  SCO_frame,
  aes(
  x = urban_centric_locale,
  y = scoScore,
  color = scoScore
  )
) +
  geom_beeswarm(shape = 18, corral = "wrap", method = "hex", cex = 1.5) + # size = 3,  
  geom_text_repel(aes(x = urban_centric_locale,
                      y = scoScore,
                      label = lea_name), size = 2.75, color = "#243142",
                  max.overlaps = 8,
                  point.padding = 0.5) +
  annotate("segment", x = 0, xend = 12.5, y = 1, yend = 1, color = "#A7A9AB",
           linetype = "dotted") +
  #scale_color_manual(values = ucl_palette) +
  scale_color_steps2(
    low = "#FFF4C6",
    high = "#990000",
    mid = "#FFAA00",
    midpoint = 1.1,
    breaks = c(0, 0.5, 0.75, 1, 1.25, 1.5)
  ) +
  ylim(0, 2) +
  ylab("School Corporation Opportunity Score") +
  xlab("NCES Urban-Centric Locale Category") +
  theme_pubr() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "none"
  )
urcl_swarm
```

We can then look at the *relative opportunity* for the various urban-centered locale
classification (@fig-lolipop). The line in the center of the plot is the *mean relative opportunity.*
As a class, school corporations in cities have higher relative opportunity than in
other locales. This does not mean that there are not high opportunity school
corporations in other locale classifications (there are), but rather this provides
a sense of high-level priority which can then be broken down further and examined
locally and in context.

```{r sco-lolipop}
#| label: fig-lolipop
#| fig-cap: "Plot of relative SCO scores by urban-centered locale classification."

sco_lol <- SCO_frame |>
  mutate(weight = enrollment * scoScore) |>
  group_by(urban_centric_locale) |>
  summarize(mean_weight = mean(weight)) |>
  ungroup() |>
  na.omit()
mean_of_means <- mean(sco_lol$mean_weight)
sco_lol$difference <- log2(sco_lol$mean_weight / mean_of_means)

sco_lol$difference[sco_lol$difference > 3] <- 3
sco_lol$difference[sco_lol$difference < -3] <- -3

sd_difference <- sd(sco_lol$difference)
low_sd <- 0 - sd_difference

compare.graph2 <-
  ggdotchart(
    sco_lol,
    x = "urban_centric_locale",
    y = "difference",
    color = "difference",
    sorting = "descending",
    add = "segments",
    rotate = TRUE,
    dot.size = 6,
    add.params = list(color = "#A7A9AB"),
    #palette = iu.colors,
    label = abs(round(sco_lol$difference, 1)),
    font.label = list(size = 8, vjust = 0.5, color = "#A7A9AB"),
    ggtheme = theme_pubr()
  ) +
  geom_hline(yintercept = 0,
             #linetype = 3,
             color = "#A7A9AB") +
  geom_hline(yintercept = sd_difference,
             linetype = 3,
             color = "#A7A9AB") +
    geom_hline(yintercept = low_sd,
             linetype = 3,
             color = "#A7A9AB") +
  scale_y_continuous(
    breaks = c(-3, low_sd, 0, sd_difference, 3),
    limits = c(-3, 3),
    labels = c("Low", "1 SD", " ", "1 SD", "High")
  ) + xlab("NCES Urban-Centric Locale Category") +
  ylab("Relative Opportunity (log2)") +
      scale_color_gradient2(
    name = "SCOscore",
    low = "#FFF4C6",
    high = "#990000",
    mid = "#FFAA00",
    midpoint = 0
  ) +
  theme(text = element_text(size = 10), legend.position = "none")
ggsave(here("outputs", "relative_opp.png"), compare.graph2, dpi = 300, width = 16, height = 9, units = "in")
compare.graph2
```

## Repositories {.appendix}

Repositories that connect with this project can be found in the following locations.

* {{< ai osf >}} [OSF Repository](https://osf.io/zt9ux/?view_only=f7d37922696b4627b9081edf694bd96d) {{< ai open-data >}}
* {{< fa brands github >}} [Github Repository](https://github.com/jeremyfprice/scoscore) {{< ai open-materials >}}

## Session Information {.appendix}

Session information is provided for reproducibility purposes.

```{r print-session-info, purl=FALSE}
sessioninfo::session_info(pkgs = "attached")
```
